---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Loading and preprocessing the data

The first part of the project requires the data to be fetched and processed 

```{r Datafetch}
## reading activity data from zipped folder 

File_name <- unzip("activity.zip", list=TRUE)
Activity_Data <- read.table(unz("activity.zip", File_name$Name),header=T, quote="\"", sep=",")

```

##  What is mean total number of steps taken per day?

Includes the required histogram 


```{r Histogram, echo=TRUE}
daywise_data <- tapply(Activity_Data$steps,as.Date(Activity_Data$date,"%Y-%m-%d"),sum)
hist(daywise_data,xlab = "Total number of steps taken each day")
```

Also the mean and median calculations day wise 


```{r mean_med, echo=TRUE}
Act_Data_mean <- tapply(Activity_Data$steps, Activity_Data$date, mean)
Act_Data_med  <- tapply(Activity_Data$steps, Activity_Data$date, median)

```

Value of mean 
```{r printm, echo=TRUE}
print (Act_Data_mean)
```
value of median 
```{r printmed, echo=TRUE}
print (Act_Data_med)
```

## What is the average daily activity pattern?

```{r drawplot, echo=TRUE}
Activity_Data_compl <- subset(Activity_Data,complete.cases(Activity_Data))
mean_steps <- tapply(Activity_Data_compl$steps, Activity_Data_compl$interval, mean)

plot(names(mean_steps),mean_steps,type = "l",xlab = "Interval",ylab = "No of Steps")
```

Maximum Number of steps 
```{r printmax, echo=TRUE}
## 5 min Interval with max steps 
Max_Mean_steps <- mean_steps[mean_steps== max(mean_steps)]
print(Max_Mean_steps)
```

Drawing the same on the plot i.e the Interval with max value of steps 
```{r drawabline, echo=TRUE}
Activity_Data_compl <- subset(Activity_Data,complete.cases(Activity_Data))
mean_steps <- tapply(Activity_Data_compl$steps, Activity_Data_compl$interval, mean)

plot(names(mean_steps),mean_steps,type = "l",xlab = "Interval",ylab = "No of Steps")
Max_Mean_steps <- mean_steps[mean_steps== max(mean_steps)]
abline(v=names(Max_Mean_steps),h=Max_Mean_steps,col="Blue")

```

## Inputing missing values

1.Total Number of missing values in the Dataset 


```{r missing, echo=TRUE}
Activity_Data_compl <- subset(Activity_Data,complete.cases(Activity_Data))
Activity_Data_missing <- subset(Activity_Data,is.na(Activity_Data$steps))
Rows_With_NA <- sum(is.na(Activity_Data))

```


2. Filling NA's and creating a new data set 

```{r filldata, echo=TRUE}
##Taking out mean for the complete cases interval wise  wise 
##For the missing cases in the original data set we are replacing them with the 
##mean for the data in the same interval 

Activity_Data_compl_mean  <-tapply(Activity_Data_compl$steps, Activity_Data_compl$interval,mean)

Updated_Activity_Data <- Activity_Data

## creating the new data set with replaced NA's from the original data set.
for (i in 1 : length(Updated_Activity_Data$interval))
{
   if(is.na(Updated_Activity_Data[i,]$steps) ){
     Updated_Activity_Data[i,]$steps <- Activity_Data_compl_mean[names(Activity_Data_compl_mean) ==   Updated_Activity_Data[i,]$interval]
  
   }
 
}


```


3. Histogram for the new data with NA's filled up 

```{r newhist, echo=TRUE}
daywise_data_new <- tapply(Updated_Activity_Data$steps,as.Date(Updated_Activity_Data$date,"%Y-%m-%d"),sum)
hist(daywise_data_new,xlab = "Total number of steps taken each day(No NA's)")

```

4.New Means and Medians with the Updated Data set

```{r newmeanmed, echo=TRUE}
Updated_Act_Data_mean <- tapply(Updated_Activity_Data$steps, Updated_Activity_Data$date, mean)
Updated_Act_Data_med  <- tapply(Updated_Activity_Data$steps, Updated_Activity_Data$date, median)

print("Updated mean") 
print(Updated_Act_Data_mean)

print("Updated median")
print(Updated_Act_Data_med)
```

## Are there differences in activity patterns between weekdays and weekends?


```{r weekday , echo=TRUE}
library(lattice)

## Adding Factor Variables based on days in The data set 


Updated_Activity_Data$day  <- weekdays(as.Date(Updated_Activity_Data$date,format="%Y-%m-%d"))

weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
Updated_Activity_Data$day <- factor(Updated_Activity_Data$day %in% weekdays1, levels = c(FALSE, TRUE) ,labels=c('weekend', 'weekday'))

## Calculating mean for the time series plot
Mean_Steps_df<- aggregate(steps ~ interval + day, data = Updated_Activity_Data, mean)

## Making the Plot 

xyplot(steps ~ interval | day, data = Mean_Steps_df, type = "l", lwd = 2,
       layout = c(1, 2), xlab = "interval", ylab = "Number of steps")
```


This completes the assignment as per requirement.



